Robust Spectroscopic Inference with Imperfect Models

### 2014-12-08 tex/paper.tex
Submitted to ApJ

%version-control tagging based off of github.com/bd-j/speccal
\input{vc}

@@

\graphicspath{{figs/}}

@@

\begin{figure}[!t]
\begin{center}
  \includegraphics{figs/kernels.pdf}
  \includegraphics{figs/high2low.pdf}
  \figcaption{({\it top}) The line-of-sight velocity distribution function, $\varphi_v$, and its 
decomposition into broadening kernels.  The instrumental kernel ({\it dotted}) is treated as 
a Gaussian, the rotation kernel ({\it dashed}) is a Gaussian-like function of the projected 
rotational velocity, and the Doppler kernel ({\it solid}) is a $\delta$-function that introduces 
the radial velocity.  In this specific case, $\sigma_v = 2.9$\,km s$^{-1}$, $v\sin{i} = 5$\,km 
s$^{-1}$, and $v_z = 7$\,km s$^{-1}$, appropriate for the example in Section \ref{subsec:wasp}.
({\it bottom}) A segment of a raw, full-resolution model spectrum and its post-processed equivalent 
after convolution by $\varphi_v$ and re-sampling at the coarser resolution of the detector pixels. 
\label{fig:broadening}}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\includegraphics{figs/chebyshev.pdf}
\vspace{-0.75cm}
\figcaption{A demonstration of our treatment for residual calibration mismatches.  The observed 
spectra at the overlap of two echelle orders ({\it top}) have slightly ($\sim$1--3\%) discrepant 
shapes.  By using Chebyshev polynomials ({\it middle, top}) one can correct for that mismatch by
adjusting the data ({\it middle, bottom}); instead, in practice we equivalently distort the model
by these polynomials ({\it bottom}) such that the model remains linear in the Chebyshev coefficients
(Eq.~\ref{eqn:chebyshev}).  \label{fig:chebyshev}}
\end{figure}

@@

\begin{figure*}[!htb]
\begin{center}
\includegraphics[scale=1.05]{figs/matrix_compilation.pdf}
\figcaption{A decomposition of the modeling procedure, explicitly highlighting the roles of the 
various contributions to the covariance matrix.  The top panels show a typical comparison between 
the data and model spectrum, along with the associated residual spectrum.  The subsequent panels 
focus on the illustrative region shaded in grey.  The left column of panels show the corresponding 
region of the covariance matrix $\vC$, decomposed into its primary contributions: from top to 
bottom, the trivial noise matrix, then combined with the global covariance kernel, and finally 
including an appropriate local covariance kernel.  In the right column, we show the zoomed-in 
residual spectrum ({\it black}) along with example random draws from the subsets of $\vC$ exhibited 
to the left.  The shaded contours ({\it orange}) represent the 1, 2, and 3\,$\sigma$ dispersions of 
an ensemble of 200 random draws from $\vC$.  Note that the trivial noise matrix ($\delta_{ij} 
\sigma_i$) poorly reproduces both the scale and structure of the residual spectrum.  The addition 
of a global kernel ($\mathcal{K}^\textrm{G}$) more closely approximates the structure and amplitude of 
the residuals, but misses the outlier line at 5202.2\,\AA.  Including a local kernel 
($\mathcal{K}^\textrm{L}$) at that location results in a covariance structure that does an excellent job 
reproducing all the key residual features.  \label{fig:matrix}}
\end{center}
\end{figure*}

@@

\begin{figure}[!t]
\begin{center}
\includegraphics{figs/badlines.pdf}
\figcaption{A particularly illustrative spectral region with substantial localized structure in 
the residuals due to ``outlier" spectral lines in the model library.  For any specific line, there 
might exist a set of model parameters, $\vT$, that will improve its match with the data, but a 
$\vT$ that will properly fit \emph{all} the outlier lines does not exist in a pre-computed library 
with (necessarily) limited parametric flexibility.  Out of concern that such intrinsic mismatches 
can bias the inference on $\vT$, the methodology advocated here introduces local kernels to inflate 
the covariance around these outliers, self-consistently down-weighting their influence on the fit.  
\label{fig:badlines}}
\end{center}
\end{figure}

@@

\begin{figure}[!b]
  \includegraphics{figs/PGM.pdf}
  \figcaption{A probabilistic graphical model representing how the parameters of the model are
    combined to forward model a spectrum.  When interpolating models using a spectral emulator
    (Appendix \ref{sec:Appendix}), the stellar parameters ($\vt_\ast$) determine the weights
    ($\mathbf{w}$) of the eigenspectra ($\mathbf{\Xi}$), which are modified according to the
    observational parameters ($\vt_\textrm{obs}$) and polynomial parameters ($\cheb$). Together,
    these parameters specify the model spectrum ($\vM$). If one uses linear interpolation instead of
    a spectral emulator, then there would be no intermediate nodes for $\mathbf{w}$ and $\mathbf{\Xi}$. The structure
    of the covariance matrix, which is included in the likelihood function, is determined by the
    covariance hyperparameters ($\cov$). Together, the model spectrum and the covariance matrix
    predict the resulting dataset ($\vD$).
\label{fig:PGM}}
\end{figure}

@@

\begin{figure*}[!t]
\includegraphics{figs/residuals_Kurucz_logg.pdf}
\vspace{-0.3cm}
\figcaption{({\it top}) A representative segment of the TRES spectrum of WASP-14 ({\it blue}), 
overlaid with a {\sc CfA/Kurucz} model ({\it red}) generated by drawing parameters from the 
posterior distribution (for a fixed $\log g = 4.29$).  ({\it bottom}) The residual spectrum along 
with contours representing the distributions of a large number of random draws from the covariance 
matrix (the shading is representative of the 1, 2, and 3\,$\sigma$ spreads of that distribution of 
draws), as in Fig.~\ref{fig:matrix}.  With such a finely tuned model library, local covariance 
kernels are not required.  \label{fig:Kurucz_residuals}}
\vspace{0.2cm}
\end{figure*}

@@

\begin{figure}[!b]
  \includegraphics{figs/Kurucz_triangle.pdf}
\vspace{-0.5cm}
  \figcaption{The posterior probability distributions for the interesting stellar parameters 
(marginalized over all other model parameters) of WASP-14 based on the {\sc CfA/Kurucz} model 
library, as explored by the MCMC Gibbs sampler.  For the two-dimensional posterior projections, 
contours are drawn at the standard 1, 2, and 3\,$\sigma$ levels (of a normal distribution) for 
reference.  \label{fig:Kurucz_posterior} }
\end{figure}

@@

\begin{figure*}[!htb]
\includegraphics{figs/residuals_PHOENIX_logg.pdf}
\vspace{-0.3cm}
\figcaption{Same as Fig.~\ref{fig:Kurucz_residuals} for the {\sc Phoenix} models.  Note the
increased vertical scale and local covariance structure for the residuals.  
\label{fig:PHOENIX_residuals}}
\vspace{0.2cm}
\end{figure*}

\begin{figure}[!b]
\begin{center}
  \includegraphics[width=0.5\textwidth]{figs/PHOENIX_triangle.pdf}
\vspace{-0.5cm}
\figcaption{Same as Fig.~\ref{fig:Kurucz_posterior}, but for the {\sc Phoenix} models.
\label{fig:PHOENIX_posterior}}
\end{center}
\end{figure}

@@

\begin{figure*}[!htb]
  \includegraphics{figs/residuals_Gl51_logg.pdf}
\vspace{-0.3cm}
  \figcaption{The $K$-band SPEX spectrum of Gl~51 ({\it blue}) compared with a
  {\sc Phoenix} model ({\it red}) generated by drawing parameters from the
  posterior distribution. ({\it bottom}) The residual spectrum along with
  contours representing the distributions of a large number of random draws
  from the covariance matrix (the shading is representative of the 1,
2, and 3\,$\sigma$ spreads of that distribution of draws), as in Fig.~\ref{fig:Kurucz_residuals}.  
Note how the ``outlier" features are clearly identified and treated by the local covariance kernels.
\label{fig:Gl51_residuals}}
\vspace{0.2cm}
\end{figure*}

\begin{figure}[!b]
  \includegraphics[width=0.5\textwidth]{figs/Gl51_triangle.pdf}
  \figcaption{The posterior probability distributions for the interesting stellar parameters
(marginalized over all other model parameters) of Gl~51 based on the {\sc Phoenix} model library,
as explored by the MCMC Gibbs sampler.  For the two-dimensional posterior projections, contours are
drawn at the standard 1, 2, and 3\,$\sigma$ levels (of a normal distribution) for reference.
\label{fig:Gl51_posterior}}
\end{figure}

@@

\acknowledgments  The authors would like to acknowledge the following people
for many extraordinarily helpful discussions and key insights: Daniel
Foreman-Mackey, Guillermo Torres, David Latham, John Johnson and the ExoLab,
Daniel Eisenstein, Rebekah Dawson, Tom Loredo and the ExoStat group, Allyson
Bieryla, and Maxwell Moe.  IC was supported by the NSF Graduate Fellowship and
the Smithsonian Institution.  Figure~\ref{fig:PGM} was generated with
\texttt{Daft}\footnote{\url{http://daft-pgm.org/}} and Figures
\ref{fig:Kurucz_posterior}, \ref{fig:PHOENIX_posterior}, and
\ref{fig:Gl51_posterior} were generated with \texttt{triangle.py}
\citep{foreman-mackey14}.  This research made extensive use of Astropy
\citep{astropy13} and the Julia programming language \citep{julia12}.

@@

\begin{figure*}[!htb]
  \begin{center}
  \includegraphics{figs/pca_reconstruct.pdf}
  \vspace{-0.6cm}
  \figcaption{ (\emph{top}) The mean spectrum, standard deviation spectrum, and five eigenspectra 
that form the basis of the \textsc{Phoenix} synthetic library used to model Gl~51, generated using 
a subset of the parameter-space most relevant for M dwarfs.  (\emph{bottom}) The original synthetic 
spectrum from the \textsc{Phoenix} library ($\vt_\ast = [T_\textrm{eff} = 3000\,\textrm{K}, \log g =
  5.0\,\textrm{dex}, \Z = 0.0\, \textrm{dex}]$) compared with a spectrum reconstructed from a linear 
combination of the derived eigenspectra using Eqn~\ref{eqn:reconstruct} (with the weights $w_k$ 
listed in the top panel figure).  \label{fig:pca_reconstruct}}
  \end{center}
\end{figure*}

@@

\begin{figure*}[!t]
\begin{center}
  \includegraphics{figs/GP_left.pdf}
  \includegraphics{figs/GP_right.pdf}
  \figcaption{The Gaussian process modelling of the principal component weights for the Gl51
    \textsc{Phoenix} spectral library. (\emph{left}) the blue dots mark the weights
    $\mathbf{w}_5^\textrm{grid}$ of the 5th eigenspectrum $\xi_5$ computed at a one-dimensional
    slice of the spectral library for grid points with $\log g = 5.0$, $\Z = 0.0$ and various values
    of $T_\textrm{eff}$. In reality, the weights are a three-dimensional function of $\vt_\ast$. The
    thin blue lines show 50 random draws of possible functional forms described by the Gaussian
    process, constrained to pass through the grid weights. (\emph{inset}) a zoomed portion showing
    the scatter in the possible functional forms. The black vertical line represents a slice through
    the scatter of the predicted weight value at $\vt_\ast = [T_\textrm{eff} = 3150\,\textrm{K},
    \log g = 5.0\,\textrm{dex}, \Z = 0.0\, \textrm{dex}]$. (\emph{right}) The posterior predictive
    probability of $w_k$ at this value of $\vt_\ast$ is completely described by
    Eqn~\ref{eqn:conditional_gp}, allowing us to analytically marginalize over all probable values
    of the weights, and thus marginalize over all probable spectral interpolations.
\label{fig:GP_interp}}
\end{center}
\end{figure*}

>>>>

@@

% it was great while it lasted
%\graphicspath{{figs/}}

@@

\begin{figure}[!t]
\begin{center}
  \includegraphics{kernels.pdf}
  \includegraphics{high2low.pdf}
  \figcaption{({\it top}) The line-of-sight velocity distribution function, $\varphi_v$, and its 
decomposition into broadening kernels.  The instrumental kernel ({\it dotted}) is treated as 
a Gaussian, the rotation kernel ({\it dashed}) is a Gaussian-like function of the projected 
rotational velocity, and the Doppler kernel ({\it solid}) is a $\delta$-function that introduces 
the radial velocity.  In this specific case, $\sigma_v = 2.9$\,km s$^{-1}$, $v\sin{i} = 5$\,km 
s$^{-1}$, and $v_z = 7$\,km s$^{-1}$, appropriate for the example in Section \ref{subsec:wasp}.
({\it bottom}) A segment of a raw, full-resolution model spectrum and its post-processed equivalent 
after convolution by $\varphi_v$ and re-sampling at the coarser resolution of the detector pixels. 
\label{fig:broadening}}
\end{center}
\end{figure}

@@

\begin{figure}[t!]
\includegraphics{chebyshev.pdf}
\vspace{-0.75cm}
\figcaption{A demonstration of our treatment for residual calibration mismatches.  The observed 
spectra at the overlap of two echelle orders ({\it top}) have slightly ($\sim$1--3\%) discrepant 
shapes.  By using Chebyshev polynomials ({\it middle, top}) one can correct for that mismatch by
adjusting the data ({\it middle, bottom}); instead, in practice we equivalently distort the model
by these polynomials ({\it bottom}) such that the model remains linear in the Chebyshev coefficients
(Eq.~\ref{eqn:chebyshev}).  \label{fig:chebyshev}}
\end{figure}

@@

\begin{figure*}[!htb]
\begin{center}
\includegraphics[scale=1.05]{matrix_compilation.pdf}
\figcaption{A decomposition of the modeling procedure, explicitly highlighting the roles of the 
various contributions to the covariance matrix.  The top panels show a typical comparison between 
the data and model spectrum, along with the associated residual spectrum.  The subsequent panels 
focus on the illustrative region shaded in grey.  The left column of panels show the corresponding 
region of the covariance matrix $\vC$, decomposed into its primary contributions: from top to 
bottom, the trivial noise matrix, then combined with the global covariance kernel, and finally 
including an appropriate local covariance kernel.  In the right column, we show the zoomed-in 
residual spectrum ({\it black}) along with example random draws from the subsets of $\vC$ exhibited 
to the left.  The shaded contours ({\it orange}) represent the 1, 2, and 3\,$\sigma$ dispersions of 
an ensemble of 200 random draws from $\vC$.  Note that the trivial noise matrix ($\delta_{ij} 
\sigma_i$) poorly reproduces both the scale and structure of the residual spectrum.  The addition 
of a global kernel ($\mathcal{K}^\textrm{G}$) more closely approximates the structure and amplitude of 
the residuals, but misses the outlier line at 5202.2\,\AA.  Including a local kernel 
($\mathcal{K}^\textrm{L}$) at that location results in a covariance structure that does an excellent job 
reproducing all the key residual features.  \label{fig:matrix}}
\end{center}
\end{figure*}

@@

\begin{figure}[!t]
\begin{center}
\includegraphics{badlines.pdf}
\figcaption{A particularly illustrative spectral region with substantial localized structure in 
the residuals due to ``outlier" spectral lines in the model library.  For any specific line, there 
might exist a set of model parameters, $\vT$, that will improve its match with the data, but a 
$\vT$ that will properly fit \emph{all} the outlier lines does not exist in a pre-computed library 
with (necessarily) limited parametric flexibility.  Out of concern that such intrinsic mismatches 
can bias the inference on $\vT$, the methodology advocated here introduces local kernels to inflate 
the covariance around these outliers, self-consistently down-weighting their influence on the fit.  
\label{fig:badlines}}
\end{center}
\end{figure}

@@

\begin{figure}[!b]
  \includegraphics{PGM.pdf}
  \figcaption{A probabilistic graphical model representing how the parameters of the model are
    combined to forward model a spectrum.  When interpolating models using a spectral emulator
    (Appendix \ref{sec:Appendix}), the stellar parameters ($\vt_\ast$) determine the weights
    ($\mathbf{w}$) of the eigenspectra ($\mathbf{\Xi}$), which are modified according to the
    observational parameters ($\vt_\textrm{obs}$) and polynomial parameters ($\cheb$). Together,
    these parameters specify the model spectrum ($\vM$). If one uses linear interpolation instead of
    a spectral emulator, then there would be no intermediate nodes for $\mathbf{w}$ and $\mathbf{\Xi}$. The structure
    of the covariance matrix, which is included in the likelihood function, is determined by the
    covariance hyperparameters ($\cov$). Together, the model spectrum and the covariance matrix
    predict the resulting dataset ($\vD$).
\label{fig:PGM}}
\end{figure}

@@

\begin{figure*}[!t]
\includegraphics{residuals_Kurucz_logg.pdf}
\vspace{-0.3cm}
\figcaption{({\it top}) A representative segment of the TRES spectrum of WASP-14 ({\it blue}), 
overlaid with a {\sc CfA/Kurucz} model ({\it red}) generated by drawing parameters from the 
posterior distribution (for a fixed $\log g = 4.29$).  ({\it bottom}) The residual spectrum along 
with contours representing the distributions of a large number of random draws from the covariance 
matrix (the shading is representative of the 1, 2, and 3\,$\sigma$ spreads of that distribution of 
draws), as in Fig.~\ref{fig:matrix}.  With such a finely tuned model library, local covariance 
kernels are not required.  \label{fig:Kurucz_residuals}}
\vspace{0.2cm}
\end{figure*}

@@

\begin{figure}[!b]
  \includegraphics{Kurucz_triangle.pdf}
\vspace{-0.5cm}
  \figcaption{The posterior probability distributions for the interesting stellar parameters 
(marginalized over all other model parameters) of WASP-14 based on the {\sc CfA/Kurucz} model 
library, as explored by the MCMC Gibbs sampler.  For the two-dimensional posterior projections, 
contours are drawn at the standard 1, 2, and 3\,$\sigma$ levels (of a normal distribution) for 
reference.  \label{fig:Kurucz_posterior} }
\end{figure}

@@

\begin{figure*}[!htb]
\includegraphics{residuals_PHOENIX_logg.pdf}
\vspace{-0.3cm}
\figcaption{Same as Fig.~\ref{fig:Kurucz_residuals} for the {\sc Phoenix} models.  Note the
increased vertical scale and local covariance structure for the residuals.  
\label{fig:PHOENIX_residuals}}
\vspace{0.2cm}
\end{figure*}

\begin{figure}[!b]
\begin{center}
  \includegraphics[width=0.5\textwidth]{PHOENIX_triangle.pdf}
\vspace{-0.5cm}
\figcaption{Same as Fig.~\ref{fig:Kurucz_posterior}, but for the {\sc Phoenix} models.
\label{fig:PHOENIX_posterior}}
\end{center}
\end{figure}

@@

\begin{figure*}[!htb]
  \includegraphics{residuals_Gl51_logg.pdf}
\vspace{-0.3cm}
  \figcaption{The $K$-band SPEX spectrum of Gl~51 ({\it blue}) compared with a
  {\sc Phoenix} model ({\it red}) generated by drawing parameters from the
  posterior distribution. ({\it bottom}) The residual spectrum along with
  contours representing the distributions of a large number of random draws
  from the covariance matrix (the shading is representative of the 1,
2, and 3\,$\sigma$ spreads of that distribution of draws), as in Fig.~\ref{fig:Kurucz_residuals}.  
Note how the ``outlier" features are clearly identified and treated by the local covariance kernels.
\label{fig:Gl51_residuals}}
\vspace{0.2cm}
\end{figure*}

\begin{figure}[!b]
  \includegraphics[width=0.5\textwidth]{Gl51_triangle.pdf}
  \figcaption{The posterior probability distributions for the interesting stellar parameters
(marginalized over all other model parameters) of Gl~51 based on the {\sc Phoenix} model library,
as explored by the MCMC Gibbs sampler.  For the two-dimensional posterior projections, contours are
drawn at the standard 1, 2, and 3\,$\sigma$ levels (of a normal distribution) for reference.
\label{fig:Gl51_posterior}}
\end{figure}

@@

\acknowledgments  The authors would like to acknowledge the following people for many
extraordinarily helpful discussions and key insights: Daniel Foreman-Mackey, Guillermo Torres, David
Latham, John Johnson and the ExoLab, Daniel Eisenstein, Rebekah Dawson, Tom Loredo and the ExoStat
group, Allyson Bieryla, and Maxwell Moe.  IC is supported by the NSF Graduate Fellowship and the
Smithsonian Institution. K.M. is supported at Harvard by NSF grant AST-1211196. Figure~\ref{fig:PGM} was generated with
\texttt{Daft}\footnote{\url{http://daft-pgm.org/}} and Figures \ref{fig:Kurucz_posterior},
\ref{fig:PHOENIX_posterior}, and \ref{fig:Gl51_posterior} were generated with \texttt{triangle.py}
\citep{foreman-mackey14}.  This research made extensive use of Astropy \citep{astropy13} and the
Julia programming language \citep{julia12}.

@@

\begin{figure*}[!htb]
  \begin{center}
  \includegraphics{pca_reconstruct.pdf}
  \vspace{-0.6cm}
  \figcaption{ (\emph{top}) The mean spectrum, standard deviation spectrum, and five eigenspectra 
that form the basis of the \textsc{Phoenix} synthetic library used to model Gl~51, generated using 
a subset of the parameter-space most relevant for M dwarfs.  (\emph{bottom}) The original synthetic 
spectrum from the \textsc{Phoenix} library ($\vt_\ast = [T_\textrm{eff} = 3000\,\textrm{K}, \log g =
  5.0\,\textrm{dex}, \Z = 0.0\, \textrm{dex}]$) compared with a spectrum reconstructed from a linear 
combination of the derived eigenspectra using Eqn~\ref{eqn:reconstruct} (with the weights $w_k$ 
listed in the top panel figure).  \label{fig:pca_reconstruct}}
  \end{center}
\end{figure*}

@@

\begin{figure*}[!t]
\begin{center}
  \includegraphics{GP_left.pdf}
  \includegraphics{GP_right.pdf}
  \figcaption{The Gaussian process modelling of the principal component weights for the Gl51
    \textsc{Phoenix} spectral library. (\emph{left}) the blue dots mark the weights
    $\mathbf{w}_5^\textrm{grid}$ of the 5th eigenspectrum $\xi_5$ computed at a one-dimensional
    slice of the spectral library for grid points with $\log g = 5.0$, $\Z = 0.0$ and various values
    of $T_\textrm{eff}$. In reality, the weights are a three-dimensional function of $\vt_\ast$. The
    thin blue lines show 50 random draws of possible functional forms described by the Gaussian
    process, constrained to pass through the grid weights. (\emph{inset}) a zoomed portion showing
    the scatter in the possible functional forms. The black vertical line represents a slice through
    the scatter of the predicted weight value at $\vt_\ast = [T_\textrm{eff} = 3150\,\textrm{K},
    \log g = 5.0\,\textrm{dex}, \Z = 0.0\, \textrm{dex}]$. (\emph{right}) The posterior predictive
    probability of $w_k$ at this value of $\vt_\ast$ is completely described by
    Eqn~\ref{eqn:conditional_gp}, allowing us to analytically marginalize over all probable values
    of the weights, and thus marginalize over all probable spectral interpolations.
\label{fig:GP_interp}}
\end{center}
\end{figure*}
